# Calculating SIMD {#simd}

For the final calculation of the SIMD score and ranks we follow a few simple steps, see the script `openSIMD_analysis/scripts/calculations/openSIMD.R`.

First we load the packages and data from the previous domain rank calculations.

```{r}
library(dplyr)
source("../openSIMD_analysis/scripts/utils/helpers.R")
domains <- read.csv("../openSIMD_analysis/results/domain_ranks.csv")
```

Then we inverse rank the domain ranks (so that low values are most deprived) and exponentially transform, spreading out the most deprived areas for a finer distinction in this region.

```{r}
invRank <- function(v) rank(-v)

exponential_domains <- domains %>%
  mutate_at(vars(-data_zone), funs(invRank)) %>%
  mutate_at(vars(-data_zone), funs(expoTransform))
```

We then combine the domain ranks via a weighting and rank the resulting combination.

```{r}
with(exponential_domains, {
  simd_score <<-
    .28 * income +
    .28 * employment +
    .14 * health +
    .14 * education +
    .09 * access +
    .05 * crime +
    .02 * housing
})
```

Finally we invert the rank once more giving us the final SIMD rank.

```{r}
simd_rank <- rank(-simd_score)
```

